Visualizing probability distributions across bivariate cyclic temporal granularities

### 2020-03-06 main.Rmd
cut literature review of different plot types

Choice of plots are dictated by the statistical transformations and geometric objects used for the visualization. The basic plot choice for our data structure is the one that can display distributions<!--In Section \ref{sec:measured-variable}, it is discussed that these densities could be estimated.-->. We will discuss few conventional and recent ways to plot distributions using both Kernel density estimates and descriptive statistics.

@@

### Descriptive statistics based displays

Most commonly used displays include the box plots [@Tukey1977-jx] which convey statistical features such as the median, quartile boundaries, hinges, whiskers and extreme outliers at a glance.<!--These displays are based on descriptive statistics, for example, $([n/2] + 1)/2^{\text{th}}$ and $(n + 1 - ([n/2] + 1)/2)^{\text{th}}$ order statistics are taken as estimates of the quartiles.--> Due to wide spread popularity and simplicity in implementation, a number of variations like Notched box plots [@Mcgill1978-hg] are proposed which provides alternate definitions of quantiles, whiskers, fences and outliers.<!--has box widths proportional to the number of points in the group and display confidence interval around medians aims to overcome some drawbacks of box plots.<!-- Challenge of boxplot. The standard box plot and all of it's variations are helpful to get an idea of the distribution at a glance. -->. However, for data less than 1000 observations, detailed estimates of tail behavior beyond the quartiles are not trustworthy. Also, the number of outliers is large for larger data sets.

<!-- Description and lit review of letter-value plots -->
The letter-value box plot [@Hofmann2017-sg] was designed to adjust for number of outliers proportional to the data size and display more reliable estimates of tail and are particularly useful for large data sets.<!--“outliers” in letter value plots are those observations beyond the most extreme letter value.--> The letter values are displayed till the depths where the letter values are reliable estimates of their corresponding quantiles and hence might lead to a lot of letter values being shown, leading to overload of information in one plot.

<!-- Description and lit review of quantile plots -->

Quantile plots display quantiles instead of quartiles in a traditional boxplot. These plots give more information than box plots and yet avoid clutter by enabling us to focus just on specific probabilities. While in a boxplot and letter value plots are well defined <!--an outlier is defined as a data point that is located outside the fences (“whiskers”) of the boxplot (e.g. outside 1.5 times the interquartile range above the upper quartile and below the lower quartile)-->, in quantile plots outliers are open to interpretation and not shown.

<!-- Description and lit review of density plots -->

<!-- The vase plot [@Benjamini1988-io, 1988] was a major revision from the concept of box plots where the width of box at each point is proportional to estimated density.  -->

### Kernel density based displays

Traditional ways to visualize kernel densities include violin plots [@Hintze1998-zi]. The shape of the violin represents the density estimate of the variable. <!--The more data points in a specific range, the larger the violin is for that range. -->Adding two density plots gives a symmetric plot which makes it easier to see the magnitude of the density and compare across categories, enabling easier detection of clusters or bumps within a distribution. It becomes less useful when there are too many levels or categories to display.

The summary plot [@Potter2010-qc] combines a minimal box plot with glyphs representing the first five moments and a sectioned density plot crossed with a violin plot and an overlay of a reference distribution. This suffers from the same problem as boxplots or violin plot, as it is combination of those two.

<!-- Description and lit review of density plots -->
 A Ridge line plot <!--(sometimes called Joy plot)--> shows the distribution for several groups. Distribution can be represented using density plots, all aligned to the same horizontal scale and presented with a slight overlap. Like other density based displays, these plots allow us to see multimodality in the distribution. However, these plots can be obscuring when there is overlap of distribution for two or more categories of the y-axis. Also, with lot of categories, it is difficult to compare the height of the densities across categories.

<!-- Description and lit review of HDR plots -->
The highest density region (HDR) box plot proposed by [@Hyndman1996-ft] displays a probability density region that contains points of relatively highest density. The probabilities for which the summarization is required can be chosen based on the requirement. These regions do not need to be contiguous and help identify multi-modality.

Each type of density display has different parameters, that need to be estimated given the data. Each is equipped with some benefits and challenges which should be borne in mind while using them for exploration. <!--For example, if we have too many categories in the variables mapped to facet or x-axis, the quantile plots might be useful for comparing patterns, whereas, other more involved methods of plotting are useful for studying anomalies, outlier or multimodal behavior-->

@@

The authors would like to thank the cohort at [NUMBATS](https://www.monash.edu/news/articles/team-profile-monash-business-analytics-team), Monash University for sharing their wisdom and experience of developing R packages, Dr. Peter Toscas [from Data61 CSIRO](https://data61.csiro.au/) for providing useful inputs on improving the analysis of smart meter application. The package `gravitas` was built under [Google Summer of Code, 2019](https://summerofcode.withgoogle.com/archive/). This article was created with knitr [@knitr] and R Markdown [@rmarkdown]. The project's Github repository <https://github.com/Sayani07/paper-gravitas> contains all materials required to reproduce this article.

>>>>

Choice of plots are dictated by the statistical transformations and geometric objects used for the visualization. The basic plot choice for our data structure is the one that can display distributions<!--In Section \ref{sec:measured-variable}, it is discussed that these densities could be estimated.-->. We will discuss few conventional and recent ways to plot distributions using both Kernel density estimates and descriptive statistics. Descriptive statistics based displays include box plots [@Tukey1977-jx] or different variations of it like Notched box plots [@Mcgill1978-hg]. More recent ways are the letter-value box plot [@Hofmann2017-sg] or Quantile plots which display quantiles instead of quartiles in a traditional boxplot. Kernel density based plots for displaying a distribution include violin plots [@Hintze1998-zi], summary plot [@Potter2010-qc], ridge line plots, highest density regions (HDR) box [@Hyndman1996-ft]. Each type of density display has different parameters, that need to be estimated given the data. Each is equipped with some benefits and challenges which should be borne in mind while using them for exploration.

@@

<!-- CONTENT CUT -->
<!-- ### Descriptive statistics based displays -->

<!-- Most commonly used displays include the box plots [@Tukey1977-jx] which convey statistical features such as the median, quartile boundaries, hinges, whiskers and extreme outliers at a glance.<!--These displays are based on descriptive statistics, for example, $([n/2] + 1)/2^{\text{th}}$ and $(n + 1 - ([n/2] + 1)/2)^{\text{th}}$ order statistics are taken as estimates of the quartiles. Due to wide spread popularity and simplicity in implementation, a number of variations like Notched box plots [@Mcgill1978-hg] are proposed which provides alternate definitions of quantiles, whiskers, fences and outliers.<!--has box widths proportional to the number of points in the group and display confidence interval around medians aims to overcome some drawbacks of box plots. Challenge of boxplot. The standard box plot and all of it's variations are helpful to get an idea of the distribution at a glance. However, for data less than 1000 observations, detailed estimates of tail behavior beyond the quartiles are not trustworthy. Also, the number of outliers is large for larger data sets. -->

<!-- <!-- Description and lit review of letter-value plots --> 
<!-- The letter-value box plot [@Hofmann2017-sg] was designed to adjust for number of outliers proportional to the data size and display more reliable estimates of tail and are particularly useful for large data sets.<!--“outliers” in letter value plots are those observations beyond the most extreme letter value. The letter values are displayed till the depths where the letter values are reliable estimates of their corresponding quantiles and hence might lead to a lot of letter values being shown, leading to overload of information in one plot. -->

<!-- <!-- Description and lit review of quantile plots -->

<!-- Quantile plots display quantiles instead of quartiles in a traditional boxplot. These plots give more information than box plots and yet avoid clutter by enabling us to focus just on specific probabilities. While in a boxplot and letter value plots are well defined <!--an outlier is defined as a data point that is located outside the fences (“whiskers”) of the boxplot (e.g. outside 1.5 times the interquartile range above the upper quartile and below the lower quartile), in quantile plots outliers are open to interpretation and not shown. -->

<!-- <!-- Description and lit review of density plots -->

<!-- <!-- The vase plot [@Benjamini1988-io, 1988] was a major revision from the concept of box plots where the width of box at each point is proportional to estimated density.  --> -->

<!-- ### Kernel density based displays -->

<!-- Traditional ways to visualize kernel densities include violin plots [@Hintze1998-zi]. The shape of the violin represents the density estimate of the variable. <!--The more data points in a specific range, the larger the violin is for that range. Adding two density plots gives a symmetric plot which makes it easier to see the magnitude of the density and compare across categories, enabling easier detection of clusters or bumps within a distribution. It becomes less useful when there are too many levels or categories to display. -->

<!-- The summary plot [@Potter2010-qc] combines a minimal box plot with glyphs representing the first five moments and a sectioned density plot crossed with a violin plot and an overlay of a reference distribution. This suffers from the same problem as boxplots or violin plot, as it is combination of those two. -->

<!-- <!-- Description and lit review of density plots -->
<!--  A Ridge line plot <!--(sometimes called Joy plot) shows the distribution for several groups. Distribution can be represented using density plots, all aligned to the same horizontal scale and presented with a slight overlap. Like other density based displays, these plots allow us to see multimodality in the distribution. However, these plots can be obscuring when there is overlap of distribution for two or more categories of the y-axis. Also, with lot of categories, it is difficult to compare the height of the densities across categories. -->

<!-- <!-- Description and lit review of HDR plots -->
<!-- The highest density region (HDR) box plot proposed by [@Hyndman1996-ft] displays a probability density region that contains points of relatively highest density. The probabilities for which the summarization is required can be chosen based on the requirement. These regions do not need to be contiguous and help identify multi-modality. -->

<!-- Each type of density display has different parameters, that need to be estimated given the data. Each is equipped with some benefits and challenges which should be borne in mind while using them for exploration. <!--For example, if we have too many categories in the variables mapped to facet or x-axis, the quantile plots might be useful for comparing patterns, whereas, other more involved methods of plotting are useful for studying anomalies, outlier or multimodal behavior-->

@@

The authors would like to thank the cohort at [NUMBATS](https://www.monash.edu/news/articles/team-profile-monash-business-analytics-team), Monash University for sharing their wisdom and experience of developing R packages, Dr. Peter Toscas [from Data61 CSIRO](https://data61.csiro.au/) for providing useful inputs on improving the analysis of smart meter application. The package `gravitas` was built under [Google Summer of Code, 2019](https://summerofcode.withgoogle.com/archive/). This article was created with knitr [@knitr, @R-knitr] and R Markdown [@rmarkdown, @R-rmarkdown]. The project's Github repository <https://github.com/Sayani07/paper-gravitas> contains all materials required to reproduce this article.
